Effectful Semantics for Substructural SSA

### 2023-11-15 pldi24/paper.tex
more related work

All of these papers give SSA semantics by translation -- that is,
they defined the semantics of an SSA program to be the semantics of
the image of the translation, whether CPS or ANF. There have also
been a number of papers which have studied the semantics of SSA directly.

@@

There have also been many operational semantics for SSA arising from
compiler verification efforts. \citet{barthe-compcert-ssa-2014} give
a

\TODO{}}
\TODO{\citet{buchwald-verified-ssa-2016}}
\TODO{\citet{herklotz-gsa-2023}}
\TODO{\citet{garbuzov-structural-cfg-2018}}
\TODO{\citet{vellum}}

\subparagraph{Denotational Semantics}

\TODO{\citet{pop-ssa-inout-2009}}

\paragraph{Weak memory}

>>>>

All of these papers give SSA semantics by translation -- that is, they
defined the semantics of an SSA program to be the semantics of the
image of the translation, with the idea that the simpler semantics of
the functional language can make proofs easier. The Grail language of
\citet{beringer-imp-fun} makes this idea explicit by giving both a
functional and imperative semantics to a low-level language, and
showing the conditions under which they coincide. \citet{schneider-imp-fun} prove
a similar result, and further mechanise the proof in Coq.

There have also been a number of papers which have studied the
semantics of SSA directly.

@@

There have also been many operational semantics for SSA arising from
compiler verification efforts. \citet{barthe-compcert-ssa-2014} give
an operational semantics of SSA as part of the CompCertSSA project,
and prove that the translation from three-address code into SSA is
semantics preserving. \citet{herklotz-gsa-2023} also formalise an
extension to SSA, ``gated SSA'', and give it a semantics, and give
semantics-preserving translations between GSA and ordinary SSA. Going
beyond CompCertSSA, \citet{vellum} have studied the semantics of the LLVM
IR itself.

There has been much less work on denotational semantics for SSA, or
directly on its equational theory. \citet{pop-ssa-inout-2009} give an
unusual denotational model of SSA in terms of the iteration structure
of a program, which they use to better understand the loop-closing
$\phi$-nodes found both in the gated SSA representation as well as
practical compilers such as GCC.  \citet{garbuzov-structural-cfg-2018}
exhibit a correspondence between an operational semantics for SSA, and
an operational semantics for the call-by-push-value calculus of
\citet{cbpv}. Despite working over an operational semantics, they use
the normal-form bisimulations of \citet{lassen-bisim} to derive an
equational theory for use in justifying optimisations. \todo{Their
operational semantics does contain not contain any effects, which
means that it cannot justify effect-dependent optimisations such as
eliminating redundant writes.}

We are the first paper to connect a weak memory semantics to the
semantics of SSA. In order to establish a rich equational theory,
we needed a number of features of the semantics which are sometimes
difficult to come by in weak memory semantics. First, sequencing
must be associative (i.e., $(t_1; t_2); t_3 \equiv t_1; (t_2; t_3)$).
Second, the semantics must support loops, including the equation
that a while-loop is equal to its unrolling. Finally, booleans need
to be distributive (i.e., $(A \otimes \mb{2}) \simeq (A + A)$).
